Gradual Soundness: Lessons from Static Python

### 2022-01-13 pj-7.1/paper.tex
pj: sec 3 pass

The \SP{} runtime system, formally known as Cinder, extends
CPython 3.8 in several ways to take full advantage of static types.
Cinder includes tailored bytecode instructions, virtual method tables, new
datatypes, a registry of typed modules, and a JIT compiler.

The key ingredients in Cinder may be of interest to other teams seeking to add
sound gradual typing to a widely-used language.
For example, a team of Node developers may wish to fork that runtime system
and experiment with bytecode instructions that rely on statci types.

@@

Cinder is a bytecode compiler that extends the Python instruction set.
All the standard Python instructions work similarly to Python 3.8;
in other words, an untyped Python program has the same semantics in
Cinder as in the standard interpreter.
The added instructions help either to express runtime checks or to
efficiently perform some action.
For example, the \code{FAST\_LEN} instruction quickly computes the
length of a built-in value.
%% TODO is it really undefined
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

Cinder adds virtual method tables (vtables) to typed classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.

Specifically, calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
JIT compilation can also upgrade vtable lookups to function calls.

%% related bytecodes: INVOKE_METHOD INVOKE_FUNCTION
%%  JIT can upgrade I_METH to I_FUN
%%  final classes get I_FUN
%% implementation: built on Python vectorcalls, so it's not totally new
%% beware: with vtables, method resolution is different than Python b/c args. get
%%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
%%
%% > This difference in behavior isn’t so much desired as just a consequence of making
%% > `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%% > the callable is placed on the stack, then the arguments, then there is a
%% > `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%% > means that first the callable is resolved, then the arguments. But with
%% > `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%% > itself, since this gives us opportunity to inline-cache the target of the
%% > call instead of always having to call something dynamic and unknown that’s
%% > on the stack. So that necessarily implies that first the arguments are
%% > resolved and placed on the stack, then the callable is resolved as part of
%% > the `INVOKE_METHOD`. -Carl

@@

As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
come with both a type and an implementation.
The implementation provides the same interface as the built-in,
but uses the type to reject some uses.

For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
with keys of type \code{K} and values of type \code{V}.
The implementation has three main components:
\begin{itemize}
  \item
    A constructor that uses a type and a Python dictionary
    to initialize a checked dictionary.
    The constructor ensures that all values of type \code{ChkDict[K, V]}
    begin with well-typed elements.
  \item
    Guarded update functions.
    Every operation that mutates or extends a checked dictionary must
    validate untyped arguments.
  \item
    A type tag for boundary checks.
    When a checked dictionary enters typed code from an untyped context,
    the runtime checks its key and value types for an exact match.
    For example, a value of type \code{ChkDict[K, V]} matches the type
    \code{ChkDict[str, dynamic]} only if \code{K} is type \code{str}
    and \code{V} is dynamic.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

\SP{} keeps track of typed functions and typed classes with a specialized classloader.
At runtime, the classloader keeps a registry of typed objects and whether or not they
are mutable.
The registry helps the bytecode reference objects and types via their module names.
The mutability bit lets \SP{} optimize access for immutable data while soundly allowing
patches to mutable data.
Overall, it is an essential piece of infrastructure to enable soundness and optimizations.

\subsection{Method-at-a-Time JIT}

FILL Cinder includes a JIT compiler for its bytecode.
The typical way to configure is to supply a list of functions to be JIT compiled.
Compilation works on any \SP{} function whether or not it is typed
(thus there is some incentive to use \SP{} even without adding type annotations).
For the biggest payoff, though, the JIT needs type information.

>>>>

The \SP{} runtime system, formally known as Cinder, extends
CPython 3.8 in several ways to take full advantage of static types.
Cinder includes tailored bytecode instructions, virtual method tables,
concrete datatypes, a registry of typed modules, and a JIT compiler.

These main ingredients of Cinder may be of interest to other teams seeking to
add sound gradual typing to an existing language.
For example, Node developers may wish to fork that runtime and experiment with
bytecode instructions that leverage sound static types.

@@

Cinder is a bytecode compiler that extends the Python instruction set.
All the standard Python instructions work similarly to Python 3.8;
in other words, an untyped Python program has the same behavior in
Cinder as in the standard interpreter.
The added instructions help with one of three tasks:
expressing runtime checks,
initializing concrete-typed values,
or efficiently performing a standard action.
As an example of the third kind, the \code{FAST\_LEN} instruction quickly
computes the length of a built-in value.
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

Cinder adds virtual method tables (vtables) to typed classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.
Calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
(Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
instructions.)
Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
function calls.

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
That being said, Cinder vtables are specialized to check untyped overrides
of typed methods against their expected return type (instead of using a layer of
indirection for the check).

%% beware: with vtables, method resolution is different than Python b/c args. get
%%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)

%% > This difference in behavior isn’t so much desired as just a consequence of making
%% > `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%% > the callable is placed on the stack, then the arguments, then there is a
%% > `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%% > means that first the callable is resolved, then the arguments. But with
%% > `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%% > itself, since this gives us opportunity to inline-cache the target of the
%% > call instead of always having to call something dynamic and unknown that’s
%% > on the stack. So that necessarily implies that first the arguments are
%% > resolved and placed on the stack, then the callable is resolved as part of
%% > the `INVOKE_METHOD`. -Carl

@@

The concrete versions of built-in data structures come with both a type and an
implementation.
The implementation provides the same interface as the built-in but uses a
type tag to reject certain operations.

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    First, the constructor uses a type and a Python dictionary ($\sptrawpydict$)
    to initialize a checked dictionary.
    This constructor ensures that all values of type $\sptchkdict{K}{V}$
    begin with well-typed elements.
  \item
    Second, all write operations check their input.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    Third, the dictionary value stores a type tag to support casts from the dynamic type.
    When a checked dictionary enters typed code from an untyped context,
    the runtime checks its type tag to see whether the key and value types are an exact match.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
    and $V$ is equal to $\sptdyn$,
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

\SP{} keeps track of typed functions and typed classes with a specialized classloader.
At runtime, the classloader keeps a registry of typed objects and whether or not they
are mutable.
The registry helps the bytecode reference objects and types via their module names.
The mutability bit lets \SP{} optimize access for immutable data while soundly allowing
patches to mutable data.

\subsection{Method-at-a-Time JIT}
\label{s:jit}

Cinder includes a JIT (just-in-time) compiler for its bytecode.
At the moment, programmers enable the JIT by supplying a list of method names
to the compiler.
Any method can be JIT-compiled whether or not it uses \SP{} types,
though types are needed to enable the most-effective optimizations.
